---
title: "Appendix: Simulation Codes"
author: "Maryam Loni"
date: "2026-07-08"
output: html_document
---

## 1. Simulation comparing superiority and equivalence testing (with block sensitivity)

```{r}
set.seed(123)

# -----------------------------
# 1. Design
# -----------------------------
corner_design <- expand.grid(
  A = c(-1, 1),
  B = c(-1, 1),
  C = c(-1, 1),
  D = c(-1, 1)
)
# ABCD defining-contrast blocking scheme, orthogonal to A-D
corner_design$block <- ifelse(corner_design$A * corner_design$B * corner_design$C * corner_design$D > 0, 1, 2)

center_points <- data.frame(
  A = 0, B = 0, C = 0, D = 0,
  block = rep(c(1, 2), each = 4)
)
design <- rbind(corner_design, center_points)
design$block <- factor(design$block)

# -----------------------------
# 2. Simulation settings
# -----------------------------
n_sim <- 10000
true_effects  <- c(0, 0.10, 0.20, 0.30)
sigmas        <- c(0.08, 0.10, 0.13, 0.15)
alphas        <- c(0.025, 0.05)
block_effects <- c(0, 0.05)          # no block effect vs. a real block effect
log_beta_0    <- log(375)
margin_practical <- log(1.25)
margin_sd        <- 0.10
results_all <- data.frame()

# -----------------------------
# 3. Simulation loop
# -----------------------------
for (true_block_eff in block_effects) {
  for (true_eff in true_effects) {
    for (log_sigma in sigmas) {
      for (alpha in alphas) {

        log_beta_C <- log(1 + true_eff)  # center-to-high effect on log scale

        detect_C <- numeric(n_sim)
        equiv_C_practical <- numeric(n_sim)
        equiv_C_sd <- numeric(n_sim)
        estimate_C <- numeric(n_sim)
        ci_low_C <- numeric(n_sim)
        ci_high_C <- numeric(n_sim)
        estimate_pct <- numeric(n_sim)
        ci_low_pct <- numeric(n_sim)
        ci_high_pct <- numeric(n_sim)
        ci_width_pct <- numeric(n_sim)
        lead_log <- numeric(n_sim)
        lead_pct <- numeric(n_sim)

        for (i in 1:n_sim) {
          # simulate response on log scale, including a true block effect
          log_Y <- log_beta_0 +
            log_beta_C * design$C +
            true_block_eff * (design$block == 2) +
            rnorm(nrow(design), mean = 0, sd = log_sigma)

          fit <- lm(log_Y ~ A + B + C + D + block, data = design)

          # superiority test for factor C
          pvals <- summary(fit)$coefficients[, 4]
          detect_C[i] <- ifelse(pvals["C"] < alpha, 1, 0)

          # equivalence: (1 - 2*alpha) CI, per TOST convention
          ci_level <- 1 - 2 * alpha
          ci <- confint(fit, level = ci_level)
          estimate_C[i] <- coef(fit)["C"]
          ci_low_C[i]   <- ci["C", 1]
          ci_high_C[i]  <- ci["C", 2]

          equiv_C_practical[i] <- ci_low_C[i] > -margin_practical & ci_high_C[i] < margin_practical
          equiv_C_sd[i]        <- ci_low_C[i] > -margin_sd & ci_high_C[i] < margin_sd

          # convert to percent scale
          estimate_pct[i] <- (exp(estimate_C[i]) - 1) * 100
          ci_low_pct[i]   <- (exp(ci_low_C[i]) - 1) * 100
          ci_high_pct[i]  <- (exp(ci_high_C[i]) - 1) * 100
          ci_width_pct[i] <- ci_high_pct[i] - ci_low_pct[i]

          # LEAD: smallest symmetric margin that would contain the CI
          lead_log[i] <- max(abs(ci_low_C[i]), abs(ci_high_C[i]))
          lead_pct[i] <- (exp(lead_log[i]) - 1) * 100
        }

        temp <- data.frame(
          true_block_effect = true_block_eff,
          true_effect_pct = true_eff * 100,
          sigma = log_sigma,
          alpha = alpha,
          superiority_rate = mean(detect_C) * 100,
          equivalence_practical = mean(equiv_C_practical) * 100,
          equivalence_sd = mean(equiv_C_sd) * 100,
          mean_estimate_log = mean(estimate_C),
          mean_ci_low_log = mean(ci_low_C),
          mean_ci_high_log = mean(ci_high_C),
          mean_ci_width_log = mean(ci_high_C - ci_low_C),
          mean_estimate_pct = mean(estimate_pct),
          mean_ci_low_pct = mean(ci_low_pct),
          mean_ci_high_pct = mean(ci_high_pct),
          mean_ci_width_pct = mean(ci_width_pct),
          mean_LEAD_log = mean(lead_log),
          mean_LEAD_pct = mean(lead_pct)
        )
        results_all <- rbind(results_all, temp)
      }
    }
  }
}

results_all
write.csv(results_all, "simulation_results_main_effects_with_block_sensitivity.csv", row.names = FALSE)
```

## 2. Sample size and power simulation

```{r}
set.seed(123)

# -----------------------------
# Fixed assumptions
# -----------------------------
target_power  <- 0.90
sigma_assumed <- 0.10     # matches observed ~10% CV at center points (Table 3)
alpha         <- 0.05
n_sim         <- 10000
log_beta_0    <- log(375)
true_effects  <- c(0, 0.05)

margins <- c(
  biologically_relevant = log(1.25),
  exploratory            = log(1.10)
)

corner_design <- expand.grid(A = c(-1, 1), B = c(-1, 1), C = c(-1, 1), D = c(-1, 1))
corner_design$block <- rep(c(1, 2), times = 8)

candidate_k <- seq(2, 60, by = 2)   # center points per block to search over

# -----------------------------
# Simulation loop
# -----------------------------
power_results <- data.frame()

for (margin_name in names(margins)) {
  margin_value <- margins[[margin_name]]

  for (true_eff in true_effects) {
    log_beta_C <- log(1 + true_eff)

    for (k in candidate_k) {
      center_points <- data.frame(A = 0, B = 0, C = 0, D = 0, block = rep(c(1, 2), each = k))
      design <- rbind(corner_design, center_points)
      design$block <- factor(design$block)

      equiv <- numeric(n_sim)
      for (i in 1:n_sim) {
        log_Y <- log_beta_0 + log_beta_C * design$C +
          rnorm(nrow(design), mean = 0, sd = sigma_assumed)
        fit <- lm(log_Y ~ A + B + C + D + block, data = design)
        ci <- confint(fit, level = 1 - 2 * alpha)
        equiv[i] <- ci["C", 1] > -margin_value & ci["C", 2] < margin_value
      }

      power_results <- rbind(power_results, data.frame(
        margin             = margin_name,
        true_effect_pct    = true_eff * 100,
        n_center_per_block = k,
        N_total            = nrow(design),
        power_pct          = mean(equiv) * 100
      ))
    }
  }
}

power_results

# Smallest sample size reaching target power, per margin x true effect
required <- do.call(rbind, lapply(
  split(power_results, list(power_results$margin, power_results$true_effect_pct)),
  function(df) {
    hit <- df[df$power_pct >= target_power * 100, ]
    if (nrow(hit) == 0) {
      data.frame(margin = df$margin[1], true_effect_pct = df$true_effect_pct[1],
                 n_center_per_block = NA, N_total = NA, power_pct = NA)
    } else {
      hit[1, ]
    }
  }
))
required

write.csv(power_results, "power_results_by_margin_and_effect.csv", row.names = FALSE)
```

## 3. Block sensitivity summary

```{r}
library(dplyr)
library(tidyr)

block_comparison <- results_all %>%
  select(true_block_effect, true_effect_pct, sigma, alpha,
         superiority_rate, equivalence_practical) %>%
  pivot_wider(
    names_from = true_block_effect,
    values_from = c(superiority_rate, equivalence_practical),
    names_prefix = "block_"
  ) %>%
  mutate(
    diff_superiority = superiority_rate_block_0.05 - superiority_rate_block_0,
    diff_equivalence = equivalence_practical_block_0.05 - equivalence_practical_block_0
  )

block_comparison
```

## 4. Result plots

The code below produces the equivalence-rate and superiority-rate plots (Figures showing decision rates by true effect, block scenario, and variability). 

```{r}
library(ggplot2)

ggplot(results_all %>% filter(alpha == 0.05),
       aes(x = true_effect_pct, y = equivalence_practical,
           color = factor(true_block_effect), linetype = factor(sigma))) +
  geom_line() +
  geom_point() +
  labs(
    x = "True effect (%)",
    y = "Equivalence rate (%)",
    color = "True block effect",
    linetype = "Sigma",
    title = "Equivalence rate by true effect, block scenario, and variability"
  ) +
  theme_bw()
```

```{r}
ggplot(results_all %>% filter(alpha == 0.05),
       aes(x = true_effect_pct, y = superiority_rate,
           color = factor(true_block_effect), linetype = factor(sigma))) +
  geom_line() +
  geom_point() +
  labs(
    x = "True effect (%)",
    y = "Superiority detection rate (%)",
    color = "True block effect",
    linetype = "Sigma",
    title = "Superiority detection rate by true effect, block scenario, and variability"
  ) +
  theme_bw()
```

### Comparison of superiority and equivalence decisions (representative version, alpha = 0.05, exploratory margin)

```{r}
library(tidyr)

results_binary <- results_all
results_binary$not_equiv_sd <- 100 - results_binary$equivalence_sd

plot_dat_sd_005 <- subset(results_binary, alpha == 0.05)[, c(
  "true_effect_pct", "sigma", "superiority_rate", "not_equiv_sd"
)]

plot_dat_sd_005 <- pivot_longer(
  plot_dat_sd_005,
  cols = c("superiority_rate", "not_equiv_sd"),
  names_to = "Decision",
  values_to = "Rate"
)

plot_dat_sd_005$Decision <- factor(
  plot_dat_sd_005$Decision,
  levels = c("superiority_rate", "not_equiv_sd"),
  labels = c("Superiority: effect detected", "Equivalence: not equivalent")
)

p_sd_005 <- ggplot(plot_dat_sd_005,
                   aes(x = true_effect_pct, y = Rate,
                       color = Decision, group = Decision)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.8) +
  facet_wrap(~ sigma, labeller = label_bquote(sigma == .(sigma))) +
  scale_x_continuous(breaks = c(0, 10, 20, 30)) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    title = "Comparison of superiority and equivalence decisions",
    subtitle = "exploratory margin (+/-0.10 on log scale), alpha = 0.05",
    x = "True effect (%)",
    y = "Decision probability (%)",
    color = "Decision"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )

p_sd_005
```

*(The same code, with `alpha == 0.025` and/or `equivalence_practical` in place of `equivalence_sd`, was used to generate the remaining alpha/margin combinations of this figure.)*

### Average confidence intervals and LEAD values (representative version, alpha = 0.025)

```{r}
plot_dat_lead_0025 <- subset(results_all, alpha == 0.025)

upper_sd_pct <- (exp(0.10) - 1) * 100
lower_sd_pct <- (exp(-0.10) - 1) * 100
upper_practical_pct <- (exp(log(1.25)) - 1) * 100
lower_practical_pct <- (exp(-log(1.25)) - 1) * 100

p_lead_0025 <- ggplot(
  plot_dat_lead_0025,
  aes(x = true_effect_pct, y = mean_estimate_pct)
) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black", linewidth = 0.7) +
  geom_hline(yintercept = lower_sd_pct, linetype = "dashed", color = "red", linewidth = 0.8) +
  geom_hline(yintercept = upper_sd_pct, linetype = "dashed", color = "red", linewidth = 0.8) +
  geom_hline(yintercept = lower_practical_pct, linetype = "dashed", color = "darkgreen", linewidth = 0.8) +
  geom_hline(yintercept = upper_practical_pct, linetype = "dashed", color = "darkgreen", linewidth = 0.8) +
  geom_errorbar(aes(ymin = mean_ci_low_pct, ymax = mean_ci_high_pct), width = 1.2, linewidth = 0.8) +
  geom_point(size = 3, color = "black") +
  geom_text(
    aes(y = mean_ci_high_pct + 3, label = paste0(round(mean_LEAD_pct, 1), "%")),
    color = "blue", size = 2.5
  ) +
  facet_wrap(~ sigma, nrow = 1, labeller = label_bquote(sigma == .(sigma))) +
  scale_x_continuous(breaks = c(0, 10, 20, 30)) +
  labs(
    title = "Average confidence intervals and LEAD values",
    subtitle = "Blue labels indicate the minimum equivalence margin required (LEAD)\nRed dashed lines: +/-0.10 log margin; green dashed lines: 0.80-1.25 practical margin",
    x = "True effect (%)",
    y = "Estimated effect and confidence interval (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    panel.grid.minor = element_blank()
  )

p_lead_0025
```

*(The same code, with `alpha == 0.05`, was used to generate the corresponding figure for alpha = 0.05.)*
