---
title: "Appendix: Case Study Codes"
author: "Maryam Loni"
date: "2026-07-08"
output: html_document
---

## 1. Data import and preparation

```{r}
dat <- read.csv2("FEC24-47-1251.csv", stringsAsFactors = FALSE)

# Convert types
dat$Std   <- as.integer(dat$Std)
dat$Run   <- as.integer(dat$Run)
dat$Block <- factor(dat$Block)   # "Block 1", "Block 2"

# Rename factor columns for clarity
names(dat)[names(dat) == "A.Coating.antibody.concentration"] <- "A_coating"
names(dat)[names(dat) == "B.Detection.antibody.concentration"] <- "B_detection"
names(dat)[names(dat) == "C.Sample.incubation.time"]          <- "C_time"
names(dat)[names(dat) == "D.Conjugate.concentration"]         <- "D_conj"

# Ensure responses are numeric (can be read as character from csv2)
dat$R1 <- as.numeric(dat$R1)
dat$R2 <- as.numeric(dat$R2)
dat$R3 <- as.numeric(dat$R3)
dat$R4 <- as.numeric(dat$R4)

str(dat)
```

## 2. Coding factor levels as -1, 0, +1

```{r}
code_factor <- function(x) {
  ux <- sort(unique(x))
  if (length(ux) != 3) stop("Factor must have exactly 3 levels")
  low  <- ux[1]
  high <- ux[3]
  ifelse(x == low, -1,
         ifelse(x == high, 1, 0))
}

dat$A_coded <- code_factor(dat$A_coating)
dat$B_coded <- code_factor(dat$B_detection)
dat$C_coded <- code_factor(dat$C_time)
dat$D_coded <- code_factor(dat$D_conj)

head(dat[, c("A_coating","A_coded",
             "B_detection","B_coded",
             "C_time","C_coded",
             "D_conj","D_coded")])
```

## 3. Equivalence summary table functions

The following two functions compute equivalence tables on the log scale: `summary_table_main_log()` for responses with main-effects-only models (R2-R4), and `summary_table_R1_log()` for R1, which uses contrast vectors to obtain the conditional effects of B and C given the B:C interaction.

```{r}
summary_table_main_log <- function(fit, response_name, Delta_practical, Delta_sd, alpha = 0.05) {
  df <- df.residual(fit)
  crit <- qt(1 - alpha, df)

  beta <- coef(fit)
  V <- vcov(fit)
  effects <- c("A_coded", "B_coded", "C_coded", "D_coded")

  out <- lapply(effects, function(term) {
    est <- beta[term]
    se  <- sqrt(V[term, term])
    ci  <- est + c(-1, 1) * crit * se
    p   <- summary(fit)$coefficients[term, 4]

    equiv_practical <- ifelse(ci[1] > -Delta_practical & ci[2] < Delta_practical,
                              "Equivalence", "Not equivalent")
    equiv_sd <- ifelse(ci[1] > -Delta_sd & ci[2] < Delta_sd,
                       "Equivalence", "Not equivalent")

    low_pct  <- (exp(ci[1]) - 1) * 100
    high_pct <- (exp(ci[2]) - 1) * 100

    data.frame(
      Response = response_name,
      Effect = gsub("_coded", "", term),
      Estimate_pct = round((exp(est) - 1) * 100, 2),
      CI_pct = paste0("(", round(low_pct, 2), ", ", round(high_pct, 2), ")"),
      P_value = round(p, 4),
      Practical_margin = equiv_practical,
      Exploratory_margin = equiv_sd,
      LEAD_pct = max(abs(low_pct), abs(high_pct)),
      stringsAsFactors = FALSE
    )
  })

  do.call(rbind, out)
}
```

```{r}
summary_table_R1_log <- function(fit, contrast_list, Delta_practical, Delta_sd, alpha = 0.05) {
  df <- df.residual(fit)
  crit <- qt(1 - alpha, df)

  beta <- coef(fit)
  V <- vcov(fit)

  out <- lapply(names(contrast_list), function(nm) {
    L <- contrast_list[[nm]]

    est <- sum(L * beta)
    se  <- sqrt(as.numeric(t(L) %*% V %*% L))
    ci  <- est + c(-1, 1) * crit * se

    tval <- est / se
    p <- 2 * (1 - pt(abs(tval), df))

    equiv_practical <- ifelse(ci[1] > -Delta_practical & ci[2] < Delta_practical,
                              "Equivalence", "Not equivalent")
    equiv_sd <- ifelse(ci[1] > -Delta_sd & ci[2] < Delta_sd,
                       "Equivalence", "Not equivalent")

    low_pct  <- (exp(ci[1]) - 1) * 100
    high_pct <- (exp(ci[2]) - 1) * 100

    data.frame(
      Response = "R1",
      Effect = nm,
      Estimate_pct = round((exp(est) - 1) * 100, 2),
      CI_pct = paste0("(", round(low_pct, 2), ", ", round(high_pct, 2), ")"),
      P_value = round(p, 4),
      Practical_margin = equiv_practical,
      Exploratory_margin = equiv_sd,
      LEAD_pct = max(abs(low_pct), abs(high_pct)),
      stringsAsFactors = FALSE
    )
  })

  out <- do.call(rbind, out)

  # relabel contrast names for readability
  out$Effect[out$Effect == "A"] <- "Factor A"
  out$Effect[out$Effect == "D"] <- "Factor D"
  out$Effect[out$Effect == "C | B=-1"] <- "Factor C (at B low)"
  out$Effect[out$Effect == "C | B=0"]  <- "Factor C (at B center)"
  out$Effect[out$Effect == "C | B=+1"] <- "Factor C (at B high)"
  out$Effect[out$Effect == "B | C=-1"] <- "Factor B (at C low)"
  out$Effect[out$Effect == "B | C=0"]  <- "Factor B (at C center)"
  out$Effect[out$Effect == "B | C=+1"] <- "Factor B (at C high)"
  out$Effect[out$Effect == "BC interaction"] <- "BC"

  out
}
```

## 4. Model fitting

```{r}
fit_R1_log <- lm(log(R1) ~ Block + A_coded + B_coded + C_coded + D_coded
                 + B_coded * C_coded, data = dat)
fit_R2_log <- lm(log(R2) ~ Block + A_coded + B_coded + C_coded + D_coded, data = dat)
fit_R3_log <- lm(log(R3) ~ Block + A_coded + B_coded + C_coded + D_coded, data = dat)
fit_R4_log <- lm(log(R4) ~ Block + A_coded + B_coded + C_coded + D_coded, data = dat)
```

## 5. Contrast vectors for R1 (conditional B and C effects)

```{r}
make_L <- function(fit, ...) {
  x <- rep(0, length(coef(fit)))
  names(x) <- names(coef(fit))
  vals <- c(...)
  x[names(vals)] <- vals
  x
}

contrast_list_R1 <- list(
  "A" = make_L(fit_R1_log, A_coded = 1),
  "D" = make_L(fit_R1_log, D_coded = 1),

  "C | B=-1" = make_L(fit_R1_log, C_coded = 1, "B_coded:C_coded" = -1),
  "C | B=0"  = make_L(fit_R1_log, C_coded = 1),
  "C | B=+1" = make_L(fit_R1_log, C_coded = 1, "B_coded:C_coded" = 1),

  "B | C=-1" = make_L(fit_R1_log, B_coded = 1, "B_coded:C_coded" = -1),
  "B | C=0"  = make_L(fit_R1_log, B_coded = 1),
  "B | C=+1" = make_L(fit_R1_log, B_coded = 1, "B_coded:C_coded" = 1),

  "BC interaction" = make_L(fit_R1_log, "B_coded:C_coded" = 1)
)
```

## 6. Equivalence margins and summary tables

```{r}
Delta_practical <- log(1.25)
Delta_R1_sd <- log(1.10)
Delta_R2_sd <- log(1.10)
Delta_R3_sd <- log(1.10)
Delta_R4_sd <- log(1.10)
```

```{r}
tab_R1 <- summary_table_R1_log(
  fit = fit_R1_log, contrast_list = contrast_list_R1,
  Delta_practical = Delta_practical, Delta_sd = Delta_R1_sd, alpha = 0.05
)
tab_R2 <- summary_table_main_log(
  fit = fit_R2_log, response_name = "R2",
  Delta_practical = Delta_practical, Delta_sd = Delta_R2_sd, alpha = 0.05
)
tab_R3 <- summary_table_main_log(
  fit = fit_R3_log, response_name = "R3",
  Delta_practical = Delta_practical, Delta_sd = Delta_R3_sd, alpha = 0.05
)
tab_R4 <- summary_table_main_log(
  fit = fit_R4_log, response_name = "R4",
  Delta_practical = Delta_practical, Delta_sd = Delta_R4_sd, alpha = 0.05
)

final_summary_table <- rbind(tab_R1, tab_R2, tab_R3, tab_R4)
final_summary_table
```

```{r}
summary(fit_R1_log); summary(fit_R2_log); summary(fit_R3_log); summary(fit_R4_log)
```

## 7. Response distributions (raw and log scale)

```{r}
resp <- c("R1", "R2", "R3", "R4")

library(ggplot2)
library(dplyr)
library(tidyr)

dat_long <- dat %>%
  pivot_longer(cols = all_of(resp), names_to = "response", values_to = "value")

ggplot(dat_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 6,
                 fill = "grey85", color = "grey30") +
  geom_density(color = "steelblue", linewidth = 1) +
  facet_wrap(~response, scales = "free") +
  theme_bw() +
  labs(title = "Response distributions (raw scale)", x = NULL, y = "Density")

dat_long_log <- dat_long %>% mutate(value_log = log(value))

ggplot(dat_long_log, aes(x = value_log)) +
  geom_histogram(aes(y = after_stat(density)), bins = 6,
                 fill = "grey85", color = "grey30") +
  geom_density(color = "darkgreen", linewidth = 1) +
  facet_wrap(~response, scales = "free") +
  theme_bw() +
  labs(title = "Response distributions (log scale)", x = "log(response)", y = "Density")
```

## 8. Lack-of-fit / pure error ANOVA table

```{r}
anova_lof_table <- function(dat, response, fit, factors = c("A_coded", "B_coded", "C_coded", "D_coded")) {

  dat2 <- dat %>%
    mutate(y = log(.data[[response]])) %>%
    filter(!is.na(y))

  dat2$design_id <- interaction(dat2[, c(factors, "Block")], drop = TRUE)

  # pure error from replicated design points (center points)
  pe_tab <- dat2 %>%
    group_by(design_id) %>%
    summarise(n = n(), ybar = mean(y), ss_pe = sum((y - ybar)^2), .groups = "drop")

  SS_PE <- sum(pe_tab$ss_pe)
  df_PE <- sum(pe_tab$n - 1)

  SSE <- sum(resid(fit)^2)
  df_E <- df.residual(fit)

  SS_LOF <- SSE - SS_PE
  df_LOF <- df_E - df_PE
  MS_LOF <- SS_LOF / df_LOF
  MS_PE  <- SS_PE / df_PE
  F_LOF  <- MS_LOF / MS_PE
  p_LOF  <- 1 - pf(F_LOF, df_LOF, df_PE)

  a <- anova(fit)
  term_rows  <- rownames(a) != "Residuals"
  term_names <- rownames(a)[term_rows]

  model_ss <- sum(a[term_rows, "Sum Sq"], na.rm = TRUE)
  model_df <- sum(a[term_rows, "Df"], na.rm = TRUE)
  model_ms <- model_ss / model_df
  model_f  <- model_ms / a["Residuals", "Mean Sq"]
  model_p  <- 1 - pf(model_f, model_df, a["Residuals", "Df"])

  y_mean <- mean(dat2$y, na.rm = TRUE)
  SST <- sum((dat2$y - y_mean)^2, na.rm = TRUE)
  df_total <- nrow(dat2) - 1

  data.frame(
    Source = c("Model", term_names, "Residual", "Lack of Fit", "Pure Error", "Corrected Total"),
    `Sum of Squares` = c(model_ss, a[term_rows, "Sum Sq"], SSE, SS_LOF, SS_PE, SST),
    df = c(model_df, a[term_rows, "Df"], df_E, df_LOF, df_PE, df_total),
    `Mean Square` = c(model_ms, a[term_rows, "Sum Sq"] / a[term_rows, "Df"],
                       a["Residuals", "Mean Sq"], MS_LOF, MS_PE, NA),
    `F-value` = c(model_f, a[term_rows, "Mean Sq"] / a["Residuals", "Mean Sq"], NA, F_LOF, NA, NA),
    `p-value` = c(model_p, a[term_rows, "Pr(>F)"], NA, p_LOF, NA, NA),
    stringsAsFactors = FALSE
  )
}
```

```{r}
library(flextable)
anova_R1_log <- anova_lof_table(dat = dat, response = "R1", fit = fit_R1_log)
anova_R2_log <- anova_lof_table(dat = dat, response = "R2", fit = fit_R2_log)
anova_R3_log <- anova_lof_table(dat = dat, response = "R3", fit = fit_R3_log)
anova_R4_log <- anova_lof_table(dat = dat, response = "R4", fit = fit_R4_log)

flextable(anova_R1_log)
flextable(anova_R2_log)
flextable(anova_R3_log)
flextable(anova_R4_log)
```

## 9. Curvature test (center vs. corner points)

```{r}
curvature_test <- function(dat, response) {
  y <- log(dat[[response]])

  center_vals <- y[dat$A_coded == 0 & dat$B_coded == 0 & dat$C_coded == 0 & dat$D_coded == 0]
  corner_vals <- y[abs(dat$A_coded) == 1 & abs(dat$B_coded) == 1 & abs(dat$C_coded) == 1 & abs(dat$D_coded) == 1]

  n_c <- length(center_vals)
  n_f <- length(corner_vals)

  ybar_c <- mean(center_vals, na.rm = TRUE)
  ybar_f <- mean(corner_vals, na.rm = TRUE)

  SS_curv <- (n_f * n_c * (ybar_f - ybar_c)^2) / (n_f + n_c)

  SS_PE_center <- sum((center_vals - mean(center_vals, na.rm = TRUE))^2, na.rm = TRUE)
  df_PE_center <- length(center_vals) - 1
  MS_PE_center <- SS_PE_center / df_PE_center

  F_curv <- SS_curv / MS_PE_center
  p_curv <- 1 - pf(F_curv, 1, df_PE_center)

  data.frame(
    Response = response,
    Mean_center = ybar_c,
    Mean_corner = ybar_f,
    SS_curvature = SS_curv,
    MS_pure_error = MS_PE_center,
    df_pure_error = df_PE_center,
    F = F_curv,
    p_value = p_curv
  )
}

curv_results <- do.call(rbind, lapply(c("R1", "R2", "R3", "R4"), function(r) curvature_test(dat, r)))
curv_results
```

## 10. Coefficient of variation at center points

```{r}
center_points <- dat[dat$A_coded == 0 & dat$B_coded == 0 &
                     dat$C_coded == 0 & dat$D_coded == 0, ]

cv_lognormal <- function(x) {
  x <- x[!is.na(x) & x > 0]
  if (length(x) < 2) return(NA_real_)
  s2 <- var(log(x))
  100 * sqrt(exp(s2) - 1)
}

cv_results <- data.frame(
  Response = c("R1", "R2", "R3", "R4"),
  CV_percent = c(
    cv_lognormal(center_points$R1),
    cv_lognormal(center_points$R2),
    cv_lognormal(center_points$R3),
    cv_lognormal(center_points$R4)
  )
)

cv_results
```
